---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,results ="hide",message=FALSE,warning=FALSE,cache = TRUE)
```

## Loading and preprocessing the data
This is the code for loading and processing data into analytical one, the main procesures are:

1. Read data from a csv file;
2. Create a new variable of time from 'date' and 'intervals', thus we could know the exact time of recording;
3. Set the 'steps' count into a numeric vector for following analysis.

Here is the code:
```{r load_process, echo=TRUE}
library(dplyr)
unzip("activity.zip")
data=read.csv("activity.csv",stringsAsFactors = F)
data[,3]=sprintf("%04d", data[,3])
data=data%>%mutate(time=paste(date,interval))
data=data%>%mutate(time=as.POSIXct(strptime(time,format="%Y-%m-%d %H%M")))
data=data%>%mutate(steps=as.numeric(steps))
```


## What is mean total number of steps taken per day?
```{r mean_total_steps}
library(dplyr)
library(ggplot2)
tsteps=tapply(data$steps,data$date,mean,na.rm=T)
mtsteps=mean(tsteps,na.rm = T)
par(mar=c(5,4,1,1),las=1)
qplot(tsteps,
      geom="histogram",
      main = "Histogram for Total Steps Per Day", 
      xlab = "Total steps of one day",  
      fill=I("#56B4E9"), 
      alpha=I(.5))+
      geom_vline(xintercept=mtsteps,color="pink")
```

As illustrated in the histogram graph above, the mean total number of steps taken per day is around `r as.integer(mtsteps)`.

## What is the average daily activity pattern?

```{r average_pattern}
library(dplyr)
library(ggplot2)
# Calculate the average pattern
avgptn=tapply(data$steps,data$interval,mean,na.rm=T)
names(avgptn)=c(1:length(avgptn))
# Genarate a time series variable
times=as.POSIXct(unique(data$interval), format="%H%M")
intervals=format(as.POSIXct(strptime(times,"%Y-%m-%d %H:%M:%S",tz="")) ,format = "%H:%M")
# Locate the max steps and its time
maxsteps=max(avgptn)
maxtime=intervals[which.max(avgptn)]
# Ploting a time series graph
par(mar=c(5,4,1,1),las=1)
qplot(x=times,y=avgptn,
      geom="line",
      main="Average Daily Activity Pattern", 
      xlab="Time",
      ylab="Average Steps in 5 minutes interval",
      col=I("#56B4E9"), 
      alpha=I(.8))+
      geom_vline(xintercept=times[which.max(avgptn)],color="pink")+
      scale_x_datetime(date_labels = "%H:%M")
```

As the average steps pattern shown above, on general, our test takers walked `r round(maxsteps, digits=2)` steps in 5 minutes at `r maxtime`, which is the peak of the data.

## Imputing missing values
```{r processing_NAs}
library(dplyr)
library(ggplot2)
library(scales)

# Calculate the portion of NAs
p=mean(is.na(data$steps))
# Imput the missing values, well let's just input the mean value of each intervals into the missing data
meansp=tapply(data$steps,data$interval,mean,na.rm=T)
# Create a new data set
newdata=data
newdata[is.na(newdata$steps),1]=meansp[data[is.na(data$steps),3]]

# Ploting a time series graph
par(mar=c(5,4,1,1),las=1)
newtsteps=tapply(newdata$steps,newdata$date,mean,na.rm=T)
newmtsteps=mean(newtsteps,na.rm = T)
newmdtsteps=median(newtsteps,na.rm = T)

mix=data.frame(total_steps=c(tsteps,newtsteps),version=rep(c("original","new"),each=length(tsteps)))

par(mar=c(5,4,1,1),las=1)
qplot(data=mix,x=total_steps,geom="histogram",facets=~version,fill=version,alpha=I(.5))

# calculate the original median
mdtsteps=median(tsteps,na.rm = T)
```

`r percent(p)` of the original data are NAs. I replaced them with the average values of all non-missing records of the same interval. 

The histogram above illustrates the distribution of new data. We could find that the new data has a peak around 37 which is the mean/median of all non-missing data.

The mean value of total steps of the original data and new data are `r round(mtsteps, digits=6)` and `r round(newmtsteps, digits=6)` respectively, which are exactly the same.The median of these two data set are `r round(mdtsteps, digits=6)` and `r round(newmdtsteps, digits=6)`, which are quite close to each other under previous imputting strategy.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays_weekends,fig.width=10,fig.height=5}
library(chron)
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)
# generate weekdays/weekend data
data=data%>%mutate(wk="Weekdays")
data$wk[grepl("S(at|un)", weekdays(data$time))]="Weekends"

# calculate the average steps in each intervals within two groups (weekdays and weekends) 
wsteps=as.data.frame(tapply(data$steps,list(data$interval,data$wk),mean,na.rm=T))
wsteps=wsteps%>%mutate(time=as.POSIXct(unique(data$interval), format="%H%M"))
wsteps=gather(wsteps,key="wk",value="steps",-time)

# Ploting a time series graph
par(mar=c(5,4,1,1),mfrow=c(2,1))
qplot(data=wsteps,time,steps,color=wk,geom="line")+scale_x_datetime(date_labels = "%H:%M")+labs(color="Days")
qplot(data=wsteps,time,steps,geom="line",facets=~wk,color=wk)+scale_x_datetime(date_labels = "%H:%M")+labs(color="Days")
```

Compare two patterns in the graph above, we could see that the walking pattern in weekdays have four obvious peaks around 8:00, 12:00, 16:00 and 19:00, which shows a general moving pattern in our daily weekdays. 

In contrast, the pattern of weekends is flatter, and later than weekdays. Generally, in weekends, people tend to get up later, enjoy activities with walking more in the afternoon and night.
